---
title: 'Replication Seminar: Regression Discontinuity Analysis'
author: "Nils Droste"
date: "`r format(Sys.Date())`"
output: pdf_document
---
  
Here, we replicate Abou-Chadi, T., & Krause, W. (2018). The causal effect of radical right success on mainstream parties’ policy positions: A regression discontinuity approach. British Journal of Political Science, 1-19. https://doi.org/10.1017/S0007123418000029 

Data replication sets are available in Harvard Dataverse at https://dx.doi.org/10.7910/DVN/KYSD5S and online appendices at https://doi.org/10.1017/S0007123418000029.

# Let us prep our environment and load some functions programmed by the authors
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(magrittr)
source( here::here('Abou-ChadiKrause2018/rrp_rdd_functions.R' ))
```

Here is what I did to read transform the provided tab separated data into an RData file
```{r}
# ds <- read_tsv(here("Abou-ChadiKrause2018",'rrp_rdd.tab'))
# ds$er.in <- as.factor(ds$er.in)
# ds$er.in_l <- as.factor(ds$er.in_l)
# save(ds, file=here("Abou-ChadiKrause2018",'rrp_rdd.RData'))
```

So that we can follow their procedure (as reported in `rrp_rdd_log.pdf` or `rrp_rdd.r`)
```{r}
load( here::here('Abou-ChadiKrause2018/rrp_rdd.RData' ))
```

# Inspecting data.
  While this should have generally been done at the beginning, we will make use of the `libary(Hmisc)` package to label and describe the data.
```{r, message=FALSE, warning=FALSE}
library(Hmisc)
```

Let us create a vector of labels (from `rrp_rdd.R`)
```{r}
labelvector <- c("iso2 character country code", 
                 "election date",
                 "CMP-code mainstream party",
                 "name of mainstream party", 
                 "party family of mainstream party (CMP-coding)",
                 "electoral threshold", "electoral threshold (lagged)",
                 "radical right party vote share (centerd on electoral threshold)",
                 "radical right party vote share (centerd on electoral threshold, lagged)",
                 "radical right parliamentary presence (binary indicator)",
                 "radical right parliamentary presence (binary indicator, lagged)",
                 "rile score (according to Lowe et al. 2011)", 
                 "multiculturalism positive (CMP coding)",
                 "multiculturalism negative (CMP coding)",
                 "cultural protectionism score (Lowe et al. 2011, first difference)",
                 "per608 score (first difference)",
                 "cultural protectionism score (Kim and Fording 2003, first difference)",
                 "cultural protectionism score (Alonso and da Fonseca 2012, first difference)",
                 "cultural protectionism score (Meguid 2008, first difference)", 
                 "environment protection score (Lowe et al. 2011, first difference)") 
```

Now, let us bind variable names and labels, to print a table (using `kable()` from the `knitr` package) showingus what variables we are dealing with

```{r}
cbind(VariableName=names(ds),VariableDescription=labelvector) %>% 
  knitr::kable(caption = "Variables in Abou-Chadi & Krause (2018)")
```

Then, let us summarise the data by `summary()` (remember also `head()` and `str()` can provide us with an idea of the data)
```{r}
summary(ds)
```

Alright, so let us move on to the analysis part. First, I would want to know what vote shares of extreme right parties we are dealing with (over time).

```{r}
ds %>% ggplot(mapping = aes(y = er.v.c, x = lubridate::year(as.Date(ds$edate)))) + 
  geom_line() +
  xlab("Year") +
  ylab("Share of radical right party \n with 0 at electoral threshold") +
  facet_wrap(~iso2c)
```

Is there something you would want to know about the data? Let us try to inspect that.
Try and adapt the code below.
geom_TYPE can be `geom_plot()`, `geom_line()`, `geom_area()`, etc. check https://ggplot2.tidyverse.org/reference/
```{r}
# ds %>% ggplot(mapping = aes(y= YAXISVARIABLE, x= XAXISVARIABLE)) + geom_TYPE 
```


# Figure 1: Mainstream party position change on cultural protectionism
Nice, so now we can directly proceed to replicating the main Fig. 1

```{r}
p1 <- jump.plot( data = subset( ds , er.v.c_l <= 10 )
                 , force.var = 'er.v.c_l'
                 , yvar = 'multic.logit_fd'
                 , seat.identifier = 'er.in_l'
                 , polynomial = 1
)
p1
```

That went smoothly. Let us inspect what this `jump.plot()` function does.

```{r}
# jump.plot() # hold Strg/Cntrl and click on the function. 
# This should open another tab with the function code.
# Alternatively, we can also click on the fucntion in our Global Environment.
```


# Table 2: Mainstream party position change on cultural protectionism

Abou-Chadi and Krause also provide estimates. In particular, they provide the Local Average Treatement Effect (LATE) coeficient, which is just above and below the threshold.

```{r}
rd.multic <- rd.base( data = ds
                      , force.var = 'er.v.c_l'
                      , yvar = 'multic.logit_fd'
                      , seat.identifier = 'er.in_l'
                      , fixed.effects = 'iso2c'
                      , clust1 = 'party'
                      , clust2 = 'edate'
                      , polynomials = c( 1 , 2 , 3 , 4 )
                      , bws = NULL
)
rd.multic
```

We can certainly check  what the authors did within their `rd.base()` function. It is relatively advanced, but to cut a long story short: They built an indicator variable for the electoral threshold as we discussed in the lecture, call that indicator variable `above` and use that as an instrumental variable in `AER::ivreg(), based on  
```{r}
formula = multic.logit_fd ~ er.in_l + poly(er.v.c_l, 20, raw = TRUE) + 
    poly(force_above, 20, raw = TRUE) + as.factor(iso2c) | above + 
    poly(er.v.c_l, 20, raw = TRUE) + poly(force_above, 20, raw = TRUE) + 
    as.factor(iso2c)
```
This is an approach that is normally employed in so called fuzzy RDD designs. This confuses me a bit, not as such (it is standard for fuzzy RDD) but because in their paper they talk about using a sharp RDD approach. Anyway, here you find a somewhat simpler approach in **R**, if you want ***RDD*** code to play around for your own analysis: https://www.econometrics-with-r.org/13-4-quasi-experiments.html 

# Mainstream left and right wing party position change on cultural protectionism

After having established that there is a general LATE, Abou-Chad and Krause proceed by estimating the LATE for both left- and right-wing parties. That can be called estimating *heterogeneous treatment* effects, as the LATE may differ across groups within the treated.

## Table 3: Mainstream left party position change on cultural protectionism
```{r}
ds <- ds %>%
  dplyr::group_by( iso2c ) %>%
  dplyr::mutate( country.mean.rile.logit = mean( rile.logit , na.rm = TRUE )) %>%
  dplyr::ungroup( ) %>%
  dplyr::group_by( party ) %>%
  dplyr::mutate( mean.rile.logit = mean ( rile.logit , na.rm = TRUE )) %>%
  dplyr::ungroup( )
rd.ml <- rd.base( data = subset( ds , ( mean.rile.logit < country.mean.rile.logit ))
                  , force.var = 'er.v.c_l'
                  , yvar = 'multic.logit_fd'
                  , seat.identifier = 'er.in_l'
                  , fixed.effects = 'iso2c'
                  , clust1 = 'party'
                  , clust2 = 'edate'
                  , polynomials = c( 1 , 2 , 3 , 4 )
                  , bws = NULL
)
rd.ml
```

## Table 4: Mainstream right party position change on cultural protectionism
```{r}
rd.mr <- rd.base( data = subset( ds , ( mean.rile.logit > country.mean.rile.logit ))
                  , force.var = 'er.v.c_l'
                  , yvar = 'multic.logit_fd'
                  , seat.identifier = 'er.in_l'
                  , fixed.effects = 'iso2c'
                  , clust1 = 'party'
                  , clust2 = 'edate'
                  , polynomials = c( 1 , 2 , 3 , 4 )
                  , bws = NULL
)
rd.mr
```

Interesting, so right wing parties respond stronger to the presence of radical right parties' presence in parliament.

# Robustness Checks

To top up, Abou-Chadi and Krause, provide several robustness checks to show how stable their results are.

## Varying bandwith around electoral threshold

Here is how they calculated the data plotted in Figure 2.
```{r, echo=TRUE, results='hide'}
rd.sens.data <- rd.sens( data = ds
                         , force.var = 'er.v.c_l'
                         , yvar = 'multic.logit_fd'
                         , seat.identifier = 'er.in_l'
                         , fixed.effects = 'iso2c'
                         , clust1 = 'party'
                         , clust2 = 'edate'
                         , polynomials = c( 1 , 2 )
                         , bws = seq( 1.5 , 10 , .25 )
                         )
rd.sens.data
```

## Only countries with legally binding thresholds.
Figure 3: Mainstream party position change on cultural protectionism, countries with legally fixed threshold
```{r}
c.list <- c( 'AT' , 'BG' , 'CZ' , 'DE' , 'EE' , 'GR' , 'LV' , 'NL' , 'PL' , 'RO' , 'SE' , 'SI')
ds %>%
  filter( iso2c %in% c.list ) %>%
  filter( !( iso2c == 'AT' & edate < '1994-01-01' )) %>%
  filter( !( iso2c == 'GR' & edate < '1993-01-01' )) %>%
  filter( !( iso2c == 'SI' & edate < '2000-01-01' )) -> ds.fixed
p3 <- jump.plot( data = subset( ds.fixed , er.v.c_l <= 10 )
                 , force.var = 'er.v.c_l'
                 , yvar = 'multic.logit_fd'
                 , seat.identifier = 'er.in_l'
                 , polynomial = 3
)
p3
```

Table 5: Mainstream party position change on cultural protectionism, countries with legally fixed threshold
```{r}
rd.fixed <- rd.base( data = ds.fixed
                     , force.var = 'er.v.c_l'
                     , yvar = 'multic.logit_fd'
                     , seat.identifier = 'er.in_l'
                     , fixed.effects = 'iso2c'
                     , clust1 = 'party'
                     , clust2 = 'edate'
                     , polynomials = c( 1 , 2 , 3 , 4 )
                     , bws = NULL
)
rd.fixed
```

Seems, overall, and beside their advanced R-programming, Abou-Chadi and Krause provide relatively reliable results that main party positions change when a radical right party is present in parliament. 